MG7: Configurable and scalable 16S metagenomics data analysis

### 2016-07-25 docs/mg7-paper/out/paper.tex
fix weird conflict with paper sources

\title{MG7: Configurable and scalable 16S data analysis}

\author[1]{Alexey Alekhin} \affil[1]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[2]{Evdokim Kovach} \affil[2]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[3]{Marina Manrique} \affil[3]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[4]{Pablo Pareja-Tobes} \affil[4]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[5]{Eduardo Pareja} \affil[5]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[6]{Raquel Tobes} \affil[6]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[7]{Eduardo Pareja-Tobes} \affil[7]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}

\keywords{ Metagenomics, 16S, taxonomic profiling, Bio4j, Graph databases, Cloud
 computing, NGS, Genomics, big data, Microbiome, Environmental, 16S
 Database}

@@

\begin{abstract}
As part of the Cambrian explosion of omics data, metagenomics brings to
the table a specific, defining trait: its social essence. The
\emph{meta} prefix exerts its influence, with multitudes manifesting
themselves everywhere; from samples to data analysis, from actors
involved to (present and future) applications. Of these dimensions, data
analysis is where needs lay further from what current tools provide. Key
features are, among others, scalability, reproducibility, data
provenance and distribution, process identity and versioning. These are
the goals guiding our work in MG7, a 16S metagenomics data analysis
system. The basic principle is a new approach to data analysis, where
configuration, processes, or data locations are static, type-checked and
subject to the standard evolution of a well-maintained software project.
Cloud computing, in its Amazon Web Services incarnation, when coupled
with these ideas, produces a robust, safely configurable, scalable tool.
Processes, data, machine behaviors and their dependencies are expressed
using a set of libraries which bring as much as possible checking and
validation to the type level, without sacrificing expressiveness.
Together they form a toolkit for defining scalable cloud-based workflows
composed of stateless computations, with a static reproducible
specification of dependencies, behavior and wiring of all steps. The
modeling of taxonomy data is done using Bio4j, where the new paradigm of
graph databases allows for both a simple expression of taxonomic
assignment tasks and the calculation of taxa abundance values
considering the hierarchic structure of the taxonomy tree. MG7 includes
a new 16S reference database, \emph{16S-DB7}, built with a flexible and
sustainable update system, and the possibility of project-driven
personalization.
\end{abstract}

@@

During the past decade, metagenomics data analysis is growing
exponentially. Some of the reasons behind this are the increasing
throughput of massively parallel sequencing technologies (with the
derived decrease in sequencing costs), and the wide impact of
metagenomics studies \citep{oulas2015metagenomics}, especially in human
health (diagnostics, treatments, drug response or prevention)
\citep{bikel2015combining}. We should also mention what could be called
the microbiome explosion: all kind of microbiomes (gut, mouth, skin,
urinary tract, airway, milk, bladder) are now routinely sequenced in
different conditions of health and disease, or after different
treatments. The impact of microbiome analysis is also being felt in
environmental sciences \citep{ufarte2015discovery}, crop sciences, the
agrifood sector \citep{coughlan2015biotechnological}, bioenergy
\citep{yang2016discovery}, and biotechnology in general
\citep{cowan2015metagenomics} \citep{kodzius2015marine}. These new
possibilities for exploring the diversity of micro-organisms in the most
varied environments are opening new research areas, and drastically
changing the existing ones.

@@

\begin{itemize}
\tightlist
\item
  Static reproducible specification of dependencies and behavior of the
  different components using \emph{\protect\hyperlink{statika}{Statika}}
  and \emph{\protect\hyperlink{datasets}{Datasets}}
\item
  Parallelization and distributed analysis based on AWS, with on-demand
  infrastructure as the basic paradigm
\item
  Definition of complex workflows using
  \emph{\protect\hyperlink{loquat}{Loquat}}, a composable system for
  scaling/parallelizing stateless computations especially designed for
  AWS
\item
  A new approach to data analysis specification, management and
  specification based on working with it in exactly the same way as for
  a software project, together with the extensive use of compile-time
  structures and checks
\item
  Modeling of the taxonomy tree using the new paradigm of graph
  databases (Bio4j, \citep{pareja2015bio4j}). It facilitates the
  taxonomic assignment tasks and the calculation of the taxa abundance
  values considering the hierarchic structure of taxonomy tree
\item
  Exhaustive per-read taxonomic assignment using two complementary
  assignment algorithms Lowest Common Ancestor and Best BLAST Hit
\item
  Using a new 16S database of reference sequences (16S-DB7) with a
  flexible and sustainable system of updating and project-driven
  customization
\end{itemize}

@@

In this section we describe the resources and libraries developed by the
authors on top of which MG7 is built. All MG7 code is written in
\protect\hyperlink{scala}{Scala}, a hybrid object-functional programming
language. Scala was chosen based on the possibility of using certain
advanced programming styles, and Java interoperability, which let us
build on the vast number of existing Java libraries; we take advantage
of this when using Bio4j as an API for the NCBI taxonomy. It has support
for type-level programming, type-dependent types (through type members)
and singleton types, which permits a restricted form of dependent types
where types can depend essentially on values determined at compile time
(through their corresponding singleton types). Conversely, through
implicits one can retrieve the value corresponding to a singleton type.

@@

\protect\hyperlink{statika}{Statika} is a Scala library developed by
\emph{AA} and \emph{EP-T} which serves as a way of defining and
composing machine behaviors statically. The main component are
\textbf{bundles}. Each bundle declares a sequence of computations (its
behavior) which will be executed in an \textbf{environment}. A bundle
can \emph{depend} on other bundles, and when being executed by an
environment, its DAG (Directed Acyclic Graph) of dependencies is
linearized and run in sequence. In our use, bundles correspond to what
an EC2 instance should do and an environment to an AMI (Amazon Machine
Image) which prepares the basic configuration, downloads the Scala code
and runs it.

@@

\protect\hyperlink{datasets}{Datasets} is a Scala library developed by
\emph{AA} and \emph{EP-T} with the goal of being a Scala-embedded
mini-language for datasets and their locations. \textbf{Data} is
represented as type-indexed fields: keys are modeled as singleton types,
and values correspond to what could be called a denotation of the key: a
value of type \texttt{Location} tagged with the key type. Then a
\textbf{Dataset} is essentially a collection of data, which are
guaranteed statically to be different through type-level predicates,
making use of the value--type correspondence which can be established
through singleton types and implicits. A dataset location is then just a
list of locations formed by locations of each dataset key. All this is
based on what could be described as an embedding in Scala of an
extensible record system with concatenation on disjoint labels, in the
spirit of \citep{harper1990extensible} \citep{harper1991record}. For
that \emph{Datasets} uses the {[}ohnosequences/cosas{]}{[}cosas{]}
library.

@@

\protect\hyperlink{loquat}{Loquat} is a library developed by \emph{AA},
\emph{EK} and \emph{EP-T} designed for the execution of embarrassingly
parallel tasks using S3, SQS and EC2 Amazon services.

@@

We developed our own Scala-based type-safe eDSLs (embedded Domain
Specific Languages) for {[}FLASh{]}{[}flash{]} \citep{magovc2011flash}
and {[}BLAST{]}{[}blast{]} \citep{camacho2009blast} expressions and
their execution.

@@

\hypertarget{scala}{\subsection{Scala}\label{scala}}

@@

\hypertarget{statika}{\subsection{Statika}\label{statika}}

@@

\hypertarget{datasets}{\subsection{Datasets}\label{datasets}}

@@

\hypertarget{loquat}{\subsection{Loquat}\label{loquat}}

@@

\protect\hyperlink{statika}{statika} :
https://github.com/ohnosequences/statika
\protect\hyperlink{datasets}{datasets} :
https://github.com/ohnosequences/datasets
\protect\hyperlink{loquat}{loquat} :
https://github.com/ohnosequences/loquat {[}flash{]} :
https://github.com/ohnosequences/flash-api {[}blast{]} :
https://github.com/ohnosequences/blast-api {[}cosas{]} :
https://github.com/ohnosequences/cosas/ \protect\hyperlink{scala}{scala}
: http://www.scala-lang.org/

>>>>

\title{MG7: Configurable and scalable 16S metagenomics data analysis}

\author[1]{Alexey Alekhin\(^\dagger\)} \affil[1]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[2]{Evdokim Kovach\(^\dagger\)} \affil[2]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[3]{Marina Manrique} \affil[3]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[4]{Pablo Pareja-Tobes} \affil[4]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[5]{Eduardo Pareja} \affil[5]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[6]{Raquel Tobes} \affil[6]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[7]{Eduardo Pareja-Tobes} \affil[7]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}

\keywords{ Metagenomics, 16S, Bacterial diversity profile, Bio4j, Graph databases,
 Cloud computing, NGS, Genomic big data, Microbiome, Environmental, 16S
 Database}

@@

\begin{abstract}
As part of the Cambrian explosion of omics data, metagenomics brings to
the table a specific, defining trait: its social essence. The
\emph{meta} prefix exerts its influence, with multitudes manifesting
themselves everywhere; from samples to data analysis, from actors
involved to (present and future) applications. Of these dimensions, data
analysis is where needs lay further from what current tools provide. Key
features are, among others, scalability, reproducibility, data
provenance and distribution, process identity and versioning. These are
the goals guiding our work in MG7, a 16S metagenomics data analysis
system. The basic principle is a new approach to data analysis, where
configuration, processes, or data locations are static, type-checked and
subject to the standard evolution of a well-maintained software project.
Cloud computing, in its Amazon Web Services incarnation, when coupled
with these ideas, produces a robust, safely configurable, scalable tool.
Processes, data, machine behaviors and their dependencies are expressed
using a set of libraries which bring as much as possible checking and
validation to the type level, without sacrificing expressiveness.
Together they form a toolkit for defining scalable cloud-based workflows
composed of stateless computations, with a static reproducible
specification of dependencies, behavior and wiring of all steps. The
modeling of taxonomy data is done using Bio4j, where the new paradigm of
graph databases allows for both a simple expression of taxonomic
assignment tasks and the calculation of taxa abundance values
considering the hierarchic structure of the taxonomy tree. MG7 includes
a new 16S reference database, \emph{16S-DB7}, built with a flexible and
sustainable update system, and the possibility of project-driven
personalization. \(^\dagger\) The first and second authors contributed
equally to this work.
\end{abstract}

@@

During the past decade, metagenomics data analysis is growing
exponentially. Some of the reasons behind this are the increasing
throughput of massively parallel sequencing technologies (with the
derived decrease in sequencing costs), and the wide impact of
metagenomics studies \citep{oulas2015metagenomics}, especially in human
health (diagnostics, treatments, drug response or prevention)
\citep{bikel2015combining}. We should also mention what could be called
the microbiome explosion: all kind of microbiomes (gut, mouth, skin,
urinary tract, airway, milk, bladder) are now routinely sequenced in
different conditions of health and disease, or after different
treatments. The impact of Metagenomics is also being felt in
environmental sciences \citep{ufarte2015discovery}, crop sciences, the
agrifood sector \citep{coughlan2015biotechnological} and biotechnology
in general \citep{cowan2015metagenomics} \citep{kodzius2015marine}.
These new possibilities for exploring the diversity of micro-organisms
in the most varied environments are opening new research areas, and
drastically changing the existing ones.

@@

\begin{itemize}
\tightlist
\item
  Static reproducible specification of dependencies and behavior of the
  different components using \emph{Statika} and \emph{Datasets}
\item
  Parallelization and distributed analysis based on AWS, with on-demand
  infrastructure as the basic paradigm
\item
  Definition of complex workflows using \emph{Loquat}, a composable
  system for scaling/parallelizing stateless computations especially
  designed for AWS
\item
  A new approach to data analysis specification, management and
  specification based on working with it in exactly the same way as for
  a software project, together with the extensive use of compile-time
  structures and checks
\item
  Modeling of the taxonomy tree using the new paradigm of graph
  databases (Bio4j). It facilitates the taxonomic assignment tasks and
  the calculation of the taxa abundance values considering the
  hierarchic structure of taxonomy tree (cumulative values)
\item
  Exhaustive per-read taxonomic assignment using two complementary
  assignment algorithms Lowest Common Ancestor and Best BLAST Hit
\item
  Using a new 16S database of reference sequences (16S-DB7) with a
  flexible and sustainable system of updating and project-driven
  customization
\end{itemize}

@@

In this section we describe the resources and libraries developed by the
authors on top of which MG7 is built. All MG7 code is written in
\href{http://www.scala-lang.org/}{Scala}, a hybrid object-functional
programming language. Scala was chosen based on the possibility of using
certain advanced programming styles, and Java interoperability, which
let us build on the vast number of existing Java libraries; we take
advantage of this when using Bio4j as an API for the NCBI taxonomy. It
has support for type-level programming, type-dependent types (through
type members) and singleton types, which permits a restricted form of
dependent types where types can depend essentially on values determined
at compile time (through their corresponding singleton types).
Conversely, through implicits one can retrieve the value corresponding
to a singleton type.

@@

\href{https://github.com/ohnosequences/statika}{Statika} is a Scala
library developed by the first and last authors which serves as a way of
defining and composing machine behaviors statically. The main component
are \textbf{bundles}. Each bundle declares a sequence of computations
(its behavior) which will be executed in an \textbf{environment}. A
bundle can \emph{depend} on other bundles, and when being executed by an
environment, its DAG (Directed Acyclic Graph) of dependencies is
linearized and run in sequence. In our use, bundles correspond to what
an EC2 instance should do and an environment to an AMI (Amazon Machine
Image) which prepares the basic configuration, downloads the Scala code
and runs it.

@@

\href{https://github.com/ohnosequences/datasets}{Datasets} is a Scala
library developed by the first and last authors with the goal of being a
Scala-embedded mini-language for datasets and their locations.
\textbf{Data} is represented as type-indexed fields: keys are modeled as
singleton types, and values correspond to what could be called a
denotation of the key: a value of type \texttt{Location} tagged with the
key type. Then a \textbf{Dataset} is essentially a collection of data,
which are guaranteed statically to be different through type-level
predicates, making use of the value--type correspondence which can be
established through singleton types and implicits. A dataset location is
then just a list of locations formed by locations of each dataset key.
All this is based on what could be described as an embedding in Scala of
an extensible record system with concatenation on disjoint labels, in
the spirit of \citep{harper1990extensible} \citep{harper1991record}. For
that \emph{Datasets} uses
\href{https://github.com/ohnosequences/cosas/}{ohnosequences/cosas}
library.

@@

\href{https://github.com/ohnosequences/loquat}{Loquat} is a library
developed by the first, second and last authors designed for the
execution of embarrassingly parallel tasks using S3, SQS and EC2 Amazon
services.

@@

We developed our own Scala-based type-safe eDSLs (embedded Domain
Specific Languages) for
\href{https://github.com/ohnosequences/flash}{FLASh}
\citep{magovc2011flash} and
\href{https://github.com/ohnosequences/blast}{BLAST}
\citep{camacho2009blast} expressions and their execution.

@@

\subsection{Scala}\label{scala}

@@

\subsection{Statika}\label{statika}

@@

\subsection{Datasets}\label{datasets}

@@

\subsection{Loquat}\label{loquat}

@@
